Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-04-23 code/Rmd/MAYO_REPROCESSED.Rmd
Minor modifications

### Conditional Quantile Normalisation (CQN)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```

```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Decomposition of expression
Clustering of library normalised unadjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$Donor_ID)

@@

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$Donor_ID, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)

@@

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

@@

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)

@@

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.AD-BrainRegion.DiagnosisCER.CONTROL",
  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.CONTROL",
  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.AD-BrainRegion.DiagnosisTCX.CONTROL",
  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.CONTROL",
  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'AD-CONTROL'), 
                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'AD-CONTROL'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'AD-CONTROL'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

@@

# Estimate surrogate variables
SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                        MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv

@@

# Calculate correlation values of random effects
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)

@@

# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

@@

# Calculate correlation values of random effects
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)

@@

# Fit linear model using new weights and new design
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

>>>>

### Conditional Quantile Normalisation (CQN)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t

PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts 
PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
```

```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Decomposition of expression
Clustering of library normalised unadjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.zero, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$Donor_ID)

@@

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$Donor_ID, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = list()
  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$E.NA)
  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.zero)] = 0

@@

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION$expr.zero, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION$expr.zero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION$expr.zero, scale.=T, center = T)

@@

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Donor_ID)

@@

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.CONTROL-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.CONTROL-BrainRegion.DiagnosisCER.OTHER",
  "BrainRegion.DiagnosisCER.AD-BrainRegion.DiagnosisCER.OTHER",
  "BrainRegion.DiagnosisTCX.CONTROL-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.CONTROL-BrainRegion.DiagnosisTCX.OTHER",
  "BrainRegion.DiagnosisTCX.AD-BrainRegion.DiagnosisTCX.OTHER"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'CONTROL-AD'), 
                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'CONTROL-AD'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>%
                   dplyr::filter(Comparison == 'CONTROL-AD'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(BrainRegion+.~Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = list()
RESIDUAL.SVA.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E.NA)
RESIDUAL.SVA.GENE_EXPRESSION$expr.zero = RESIDUAL.SVA.GENE_EXPRESSION$expr.NA
RESIDUAL.SVA.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.SVA.GENE_EXPRESSION$expr.NA)] = 0

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

@@

# Estimate surrogate variables
SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E.zero,
                        MODEL1, MODEL0, n.sv = NUM.SV, 
                        B = 30, tol = 1e-20)$sv

@@

# Calculate correlation values of random effects
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$Donor_ID)

@@

# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E.NA)

# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(MODEL1), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```

@@

# Calculate correlation values of random effects
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$Donor_ID)

@@

# Fit linear model using new weights and new design
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$E.NA)
```
